---
title: "Modeling and prediction for movies"
output:
  rmarkdown::github_document
  # pdf_document: default
  # html_document:
  #   fig_height: 4
  #   highlight: pygments
  #   theme: spacelab
---

## Setup

```{r, echo = FALSE}
knitr::opts_chunk$set(
  fig.path = "README_figs/README-"
)
```

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(statsr)
```

### Load data

```{r load-data}
load("movies.Rdata")
```


* * *

## Part 1: Data

The data set is comprised of 651 randomly sampled movies produced and released before 2016. Because random sampling is implemented, the results can be generalized. Since there isn't random assignment, causation is not possible.  

* * *

## Part 2: Research question

We will study the IMDB rating of a movie, and try to find associations with a variety of other variables in the data set.

* * *

## Part 3: Exploratory data analysis

Let's do the first count:

```{r}
dim(movies)
```
Our dataset has 651 observations and 32 variables.

Let's count now NA values for each column:

```{r}
colSums(is.na(movies))
```
We will delete the corresponding rows from our dataset later, if necessary.

Below the description of IMDB movie rating numerical variable:

```{r}
summary(movies$imdb_rating)
ggplot(movies, aes(y = imdb_rating)) + 
  geom_boxplot() +
  ggtitle("Box Plot of IMDB rating")
```

As we can see the median is 6.6 and there are some outliers lower than 4.

Let's check now the relationships between some of the variables in the data set.

### IMDB rating and Rotten Tomatoes audience score

```{r}
ggplot(data = movies, aes(x = audience_score, y = imdb_rating)) +
  geom_jitter()
```

```{r}
movies %>%
  summarise(cor(audience_score, imdb_rating))
```

There's a quite strong linear relationship between the two variables.

### Runtime and Rotten Tomatoes audience score

```{r}
movies_clean <- filter(movies,!is.na(runtime))
ggplot(data = movies_clean, aes(x = runtime, y = audience_score)) +
  geom_jitter()
```
```{r}
movies_clean %>%
  summarise(cor(runtime, audience_score))
```
From the scatterplot and the correlation coefficient we can see a quite low positive linear correlation between the two variables.

### IMDB number of votes and Rotten Tomatoes audience score

```{r}
ggplot(data = movies, aes(x = imdb_num_votes, y = audience_score)) +
  geom_jitter()
```
```{r}
movies %>%
  summarise(cor(imdb_num_votes, audience_score))
```
In this case, as expected, there's no clear linear relationship.

### Rotten Tomatoes critics score and audience score

```{r}
ggplot(data = movies, aes(x = critics_score, y = audience_score)) +
  geom_jitter()
```
```{r}
movies %>%
  summarise(cor(critics_score, audience_score))
```
There's a quite strong positive linear correlation between the two variable. We will consider this in doing the model later.

### Categorical variables and IMDB rating

Let's plot now the relationship between the genre of a movie and the IMDB rating numerical variable:

```{r}
ggplot(data=movies,aes(x=reorder(genre,imdb_rating), y=imdb_rating)) +
  geom_boxplot() +
  labs(y="IMDB rating", x="Genre", subtitle="Boxplots of IMDB rating by movies genre") +
  theme(axis.text.x  = element_text(angle=90, vjust=0.5))
```

Below we can see the relationship between the mpaa rating and the imdb rating:

```{r}
ggplot(data=movies,aes(x=reorder(mpaa_rating,imdb_rating), y=imdb_rating)) +
  geom_boxplot() +
  labs(y="IMDB rating", x="MPAA rating", subtitle="Boxplots of IMDB rating by MPAA rating") +
  theme(axis.text.x  = element_text(angle=90, vjust=0.5))
```

* * *

## Part 4: Modeling

In this part, we will develop a multiple linear regression model to predict the **imdb_rating** variable, our response variable, using several explanatory variables.

### Choice of variables

Our response variable is **imdb_rating**.

We will exclude from our candidate explanatory variables the following variables because they are not meaningful in this context:
title, imdb_url, rt_url.

As we have seen in the previous chapter, audience_score is collinear with critics_score. We can exclude critics_score from the selected variables.

Let's analyse the relationship between imdb_num_votes and our response variable:

```{r}
movies %>%
  summarise(cor(imdb_num_votes, imdb_rating))
```
They are not so much correlated, hence we will exclude imdb_num_votes from our selected columns.

Let's analyse the relationship between runtime and our response variable:

```{r}
movies %>%
  filter(!is.na(runtime)) %>%
  summarise(cor(runtime, imdb_rating))
```
They are not so much correlated, hence we will exclude runtime from our selected columns.

As we have already seen in the previous chapter there is a quite strong positive correlation between audience score and IMDBB rating, suggesting that we could use the audience score on Rotten Tomatoes to predict the IMDB rating.

Let's create one variable named awarded with value 'yes' if one of the following variables is 'yes', 'no' otherwise. We will include this new variable in the model.

```{r}
movies_reduced <- movies %>% 
  mutate(awarded = if_else(best_pic_nom == 'yes' | best_pic_win == 'yes' | best_actor_win == 'yes' | best_actress_win == 'yes' | best_dir_win == 'yes', 'yes', 'no'))
```

We will exclude also the following variables: 
title_type, director, actor1, actor2, actor3, actor4, actor5, studio, thtr_rel_year, thtr_rel_month, thtr_rel_day, dvd_rel_year, dvd_rel_month, dvd_rel_day.

Let's select our candidates explanatory variables and our response variable:

```{r}
movies_reduced <- movies_reduced %>% select('imdb_rating','audience_score','genre','mpaa_rating','awarded','top200_box') 
```

Now let's pick one of the observations to do the prediction later:

```{r}
indices <- sample(1:nrow(movies_reduced), 1)
predict_movie <- movies_reduced[indices,]
movies_reduced <- movies_reduced[-indices,]
```


### Model selection

Our base multiple linear regression model is the following:

```{r}
movie_rating_model <- lm(imdb_rating ~ audience_score + genre + mpaa_rating + awarded + top200_box, data = movies_reduced)
m_summary <- summary(movie_rating_model)
adj_r_squared <- m_summary$adj.r.squared
adj_r_squared
```
It's a quite hight adjusted $R^2$.

Now, let's perform backward selection in order to find the best model:

```{r}
# Removing audience_score
movie_rating_model <- lm(imdb_rating ~ genre + mpaa_rating + awarded + top200_box, data = movies_reduced)
m_summary <- summary(movie_rating_model)
adj_r_squared_audience_score <- m_summary$adj.r.squared
adj_r_squared_audience_score

# Removing genre
movie_rating_model <- lm(imdb_rating ~ audience_score + mpaa_rating + awarded + top200_box, data = movies_reduced)
m_summary <- summary(movie_rating_model)
adj_r_squared_genre <- m_summary$adj.r.squared
adj_r_squared_genre

# Removing mpaa_rating
movie_rating_model <- lm(imdb_rating ~ audience_score + genre + awarded + top200_box, data = movies_reduced)
m_summary <- summary(movie_rating_model)
adj_r_squared_mpaa_rating <- m_summary$adj.r.squared
adj_r_squared_mpaa_rating

# Removing awarded
movie_rating_model <- lm(imdb_rating ~ audience_score + genre + mpaa_rating + top200_box, data = movies_reduced)
m_summary <- summary(movie_rating_model)
adj_r_squared_awarded <- m_summary$adj.r.squared
adj_r_squared_awarded

# Removing top200_box
movie_rating_model <- lm(imdb_rating ~ audience_score + genre + mpaa_rating + awarded, data = movies_reduced)
m_summary <- summary(movie_rating_model)
adj_r_squared_top200_box <- m_summary$adj.r.squared
adj_r_squared_top200_box
```
The model obtained removing mpaa_rating variable gives the best adjusted $R^2$, slightly better than the previous model with all the selected variables.

Let's remove mpaa_rating and continue backward selection until we don't see any improvement in adjusted $R^2$:

```{r}
movie_rating_model <- lm(imdb_rating ~ audience_score + genre + awarded + top200_box, data = movies_reduced)
m_summary <- summary(movie_rating_model)
adj_r_squared <- m_summary$adj.r.squared
adj_r_squared
```
Let's perform backward selection:

```{r}
# Removing audience_score
movie_rating_model <- lm(imdb_rating ~ genre + awarded + top200_box, data = movies_reduced)
m_summary <- summary(movie_rating_model)
adj_r_squared_audience_score <- m_summary$adj.r.squared
adj_r_squared_audience_score

# Removing genre
movie_rating_model <- lm(imdb_rating ~ audience_score + awarded + top200_box, data = movies_reduced)
m_summary <- summary(movie_rating_model)
adj_r_squared_genre <- m_summary$adj.r.squared
adj_r_squared_genre

# Removing awarded
movie_rating_model <- lm(imdb_rating ~ audience_score + genre + top200_box, data = movies_reduced)
m_summary <- summary(movie_rating_model)
adj_r_squared_awarded <- m_summary$adj.r.squared
adj_r_squared_awarded

# Removing top200_box
movie_rating_model <- lm(imdb_rating ~ audience_score + genre + awarded, data = movies_reduced)
m_summary <- summary(movie_rating_model)
adj_r_squared_top200_box <- m_summary$adj.r.squared
adj_r_squared_top200_box
```
No improvements, so the best model is this one:

```{r}
movie_rating_model <- lm(imdb_rating ~ audience_score + genre + awarded + top200_box, data = movies_reduced)
m_summary <- summary(movie_rating_model)
m_summary
```


### Model diagnotics

We have already analysed the linear relationship between the audience_score numerical variable and the response variable, using a scatter plot. 
Let's check the condition of nearly normal residuals with mean 0 using a histogram:

```{r}
ggplot(data = movie_rating_model, aes(x = .resid)) +
  geom_histogram(binwidth = 0.05) +
  xlab("Residuals")
```

```{r}
summary(resid(movie_rating_model))
```

As we can see the distribution is a little left skewed with mean 0.

Now, let's check the constant variability of residuals using a residuals plot:

```{r}
ggplot(data = movie_rating_model, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  xlab("Fitted values") +
  ylab("Residuals")
```

We can see that the condition of constant variability is fairly verified. Higher predicted values results in lower residuals and there is more scatter in the lower part. These need to be further analysed.

Finally, because the data are sampled randomly, the observations are independent and hence the residuals are independent.

### Interpretation of model coefficients

The p-value of the entire model is low, indicating that the model as a whole is significant. 

The p-value of the `audience_score` coefficient is low, indicating that the variable is significant. The p-value of the other categorical variables (genre, top200_box and awarded) are relative to some of the levels, so we can't say they are not entirely significant if the p-value of some levels are not low.

The estimate for the numerical variable `audience_score` is 0.04. It reflects how much higher a group of movies is expected to be rated by IMDB if they have a Rotten Tomatoes audience score that is one point higher *while holding all other variables constant*.

* * *

## Part 5: Prediction

The movie we want to predict the IMDB rating is:

```{r}
predict_movie
```


```{r}
predict(movie_rating_model, predict_movie)
```
The IMDB rating prediction for the movie (6.59) is not far from the real rating (6.9).

We can also construct a prediction interval around this prediction, which will provide a measure of uncertainty around the prediction.

```{r new-prof-predict-interval}
predict(movie_rating_model, predict_movie, interval = "prediction", level = 0.95)
```

The model predicts, with 95% confidence, that a movie with an audience score of 69, of genre Comedy, with mpaa rating PG-13, awarded, and that don't appears in the top200, is expected to have an imdb rating between 5.68 and 7.6. 

* * *

## Part 6: Conclusion

We found a significant model that explains 77.4% of the variability in the dataset. 

The predictions can be improved by adding new features not considered here, or by considering a larger sample of movies with more recent data.

