---
title: "Exploring the BRFSS data"
output: rmarkdown::github_document
---

## Setup

```{r, echo = FALSE}
knitr::opts_chunk$set(
  fig.path = "README_figs/README-"
)
```

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
```

### Load data

```{r load-data}
load("brfss2013.RData")
```

## Part 1: Data

The Behavioral Risk Factor Surveillance System (BRFSS) is a collaborative project between all of the states in the United States (US) and participating US territories and the Centers for Disease Control and Prevention (CDC). 

BRFSS is an ongoing surveillance system designed to measure behavioral risk factors for the non-institutionalized adult population (18 years of age and older) residing in the US. 

The BRFSS objective is to collect uniform, state-specific data on preventive health practices and risk behaviors that are linked to chronic diseases, injuries, and preventable infectious diseases that affect the adult population. Factors assessed by the BRFSS in 2013 include tobacco use, HIV/AIDS knowledge and prevention, exercise, immunization, health status, healthy days — health-related quality of life, health care access, inadequate sleep, hypertension awareness, cholesterol awareness, chronic health conditions, alcohol consumption, fruits and vegetables consumption, arthritis burden, and seatbelt use.

Since 2011, BRFSS conducts both landline telephone- and cellular telephone-based surveys. In conducting the BRFSS landline telephone survey, interviewers collect data from a randomly selected adult in a household. In conducting the cellular telephone version of the BRFSS questionnaire, interviewers collect data from an adult who participates by using a cellular telephone and resides in a private residence or college housing.

It is an observational study and generalizability is ensured by random sampling, but because it's not an experiment there's not causation.

## Part 2: Research questions

Following are the three research questions explored in this analysis.

**Research quesion 1:**

As first question, we are interested in exploring the relationship between the sex and the average alcoholic drink per day in past 30 days.

**Research quesion 2:**

As second question, we explore the relationship between the sex, the income level and the number of days he/she feels depressed in past 30 days.

**Research question 3:**

As third question, we are interested in the relationship between the sex, the Running or Walking activity mostly do during the last month, and the general health of the respondent.

## Part 3: Exploratory data analysis

**Research quesion 1:**

For the first question, we focus our analysis on two variables: sex of respondent, avedrnk2 (during the past 30 days, on the days when the person drank, about how many drinks did he/she drink on the average?).

```{r}
brfss2013 %>% 
  select(sex, avedrnk2) %>% 
  str()
```
Sex is a categorical variable with two values: 1 for male, 2 for female.
avedrnk2 is an numerical variable.

```{r}
brfss2013 %>% 
  group_by(sex) %>% 
  summarise(count = n())
```
Both sex and avedrnk2 have NA values, we will exclude them from our analysis.

```{r}
brfss2013 %>% 
  filter(!(is.na(sex)) & !(is.na(avedrnk2))) %>%
  group_by(sex) %>%
  summarise(mean = mean(avedrnk2), median = median(avedrnk2), sd = sd(avedrnk2), 
            min = min(avedrnk2), max = max(avedrnk2))
```
As we can see, the mean, the median and the standard deviation of average drink consumption in the past 30 days seems to be higher in male than female people.

Let's do a side by side boxplot to visualize these results:

```{r}
sex_alcohol_filtered <- brfss2013 %>%
  filter(!(is.na(sex)) & !(is.na(avedrnk2)))
ggplot(sex_alcohol_filtered, aes(x = factor(sex), y = avedrnk2)) +
  geom_boxplot()
```

**Research quesion 2:**

In this case we explore the relationship between the sex, the income level (income2 variable) and the number of days he/she feels depressed in past 30 days (qlmentl2 variable).

```{r}
brfss2013 %>%
  select(sex, income2, qlmentl2) %>% 
  str()
```
sex has two values, 1 for Male and 2 for Female. income2 has eight values. qlmentl2 is an integer variable. In all of this variables there are also some NAs that we will filter out.

Let's count:

```{r}
brfss2013 %>%
  filter(!(is.na(sex)) & !(is.na(qlmentl2))) %>%
  group_by(sex) %>% 
  summarise(count = n(), sum = sum(qlmentl2), mean = mean(qlmentl2), median = median(qlmentl2), sd = sd(qlmentl2), 
            min = min(qlmentl2), max = max(qlmentl2))
```
We can see that more women feel depressed, for more days than men on average.

Let's add the income level.

```{r}
brfss2013 %>%
  filter(!(is.na(sex)) & !(is.na(income2)) & !(is.na(qlmentl2))) %>%
  group_by(sex, income2) %>% 
  summarise(count = n(), sum = sum(qlmentl2), mean = mean(qlmentl2), median = median(qlmentl2), sd = sd(qlmentl2), 
            min = min(qlmentl2), max = max(qlmentl2))
```
It appears that, on average, female respondents spent more days feeling depressed than men. However, for a level of income of 35,000$ and more, the median of the female respondents is equal to 0 as the median of the male respondents, suggesting that for some level of income they could be the same on average.


**Research question 3:**

We explore the relationship between the sex, the Running or Walking activity mostly do during the last month (exract11 variable), and the general health of the respondent (genhlth).

```{r}
brfss2013 %>%
  select(sex, exract11, genhlth) %>% 
  str()
```
These variables are all categorical. Also in this case, we'll exclude the NA values.

First of all, let's check that Running and Walking activity are allowed values:

```{r}
levels(brfss2013$exract11)
```

Let's do some counts now. In general, how many men and women do physical activity among those listed above?

```{r}
filtered_exercise_dataset <- brfss2013 %>%
  select(sex, exract11, genhlth) %>%
  filter(!(is.na(sex)) & !(is.na(exract11)))
filtered_exercise_dataset %>%
  group_by(sex) %>%
  summarise(count = n()) %>%
  mutate(ratio=count/sum(count))
```
It appears that a little more over 58% of people that do exercise are women.

```{r}
barplot(prop.table(table(filtered_exercise_dataset$sex)))
```

Let's filter our dataset now.

```{r}
exercise_sex_health <- brfss2013 %>%
  select(sex, exract11, genhlth) %>%
  filter(!(is.na(sex)) & (exract11 == 'Running' | exract11 == 'Walking') & !(is.na(genhlth)))
exercise_sex_health %>%
  group_by(sex) %>%
  summarise(count = n()) %>%
  mutate(ratio=count/sum(count))
```
And about 62% of people that do Running or Walking activity are women. 

```{r}
barplot(prop.table(table(exercise_sex_health$sex)))
```

Let's group by sex and type of activity now:

```{r}
exercise_sex_health %>%
  group_by(sex, exract11) %>%
  summarise(count = n()) %>%
  mutate(ratio=count/sum(count))
```

Let's group by sex, type of activity and general health now:

```{r}
exercise_sex_health_grouped <- exercise_sex_health %>%
  group_by(sex, exract11, genhlth) %>%
  summarise(count = n()) %>%
  mutate(ratio = count/sum(count))
exercise_sex_health_grouped
```

Looking at the ratio by sex and type of activity, we can compare women and men general health.


